---
title: "Tutorial 2- Multidimensional Splines"
author: "Andomei Smit: SMTAND051"
date: "14/04/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
library(nlme)
library(mgcv)
library(splines)
```

# Example from the notes
```{r}
# simulate some data from a surface
n <- 5000
eg <- gamSim(2, n = n, scale = 0.5)

names(eg)
attach(eg)

truef <- eg$truth
persp(truef$x, truef$z, truef$f, col = heat.colors(40), main = "a.truth")

# linear terms plus their interaction
par(mfrow = c(1,2))
b1 <- gam(y ~ x + z + x*z, data= data)
vis.gam(b1, main = "b.linear model")

# additive splines
b2 <- gam(y ~ s(x) + s(z), data = data)
vis.gam(b2, main = "c. additive splines")


par(mfrow = c(1, 2))
ind <- sample(1:n, 200, replace = FALSE)
b5 <- gam(y ~ s(x, z, k = 40), data = data,
knots = list(x = data$x[ind], z = data$z[ind]))
vis.gam(b5, theta = 30, phi = 30, main = "low rank thin-plate spline")

## and a pure "knot based" spline of the same data
b6<- gam(y ~ s(x, z, k = 64), data = data,
knots = list(x = rep((1:8-0.5)/8,8),
z = rep((1:8-0.5)/8, rep(8,8))))
vis.gam(b6,color = "heat", theta = 30, phi = 30, main = "knot-based thin-plate spline")

summary(b6)
detach(eg)
```
# Question 1
Use the gam function to fit a tensor product spline to the surface, with and without penalty. With and without choosing the knots. Try different values of the smoothing parameter. Different numbers of knots.

## Tensor product spline with a penalty

```{r}
dat <- eg$data
y <- dat$y
x <- dat$x
z <- dat$z

tensor_penalty <- gam(y ~ te(x, z), data = dat)
vis.gam(tensor_penalty, main = "Tensor Product Spline with penalty")
```

## Tensor product spline with no penalty

```{r}
model_no_penalty <- gam(y ~ te(x, z, fx = TRUE), data = dat)
vis.gam(model_no_penalty, main = "Tensor Product Spline without penalty")
```

## Tensor Product spline with choosing number of knots manually

```{r}
range(x)
range(z)
# Penalized with 5 knots in each direction
tensor_choose_knots <- gam(y ~ te(x, z, 
                       k = c(5,5)), # set the number of knots for x and z
                data = dat)
vis.gam(tensor_choose_knots, main = "Tensor Product Spline by choosing number of knots")
```

## Tensor Product Spline choosing exact knot placement

```{r}
# Create a grid of knots
knots_list <- list(
  x = seq(0, 1, length.out = 5),
  z = seq(0, 1, length.out = 5)
)

model_custom_knots <- gam(y ~ te(x, z, k = c(5,5)), data = dat, knots = knots_list)
vis.gam(model_custom_knots, main = "Tensor Product Spline choosing exact knot placement")

```

## Tensor Product Spline with different number for smoothing parameter

```{r}
# Fix smoothing parameter to a small value (less smooth)
model_sp_small <- gam(y ~ te(x, z), data = dat, 
                      sp = c(0.01, 0.01)) # specify the smoothing function for x, z
# note that they only apply marginal penalties and do not penalize the interaction terms

# Fix smoothing parameter to a large value (very smooth)
model_sp_large <- gam(y ~ te(x, z), data = dat, sp = c(10, 10))

par(mfrow=c(1,2))
vis.gam(model_sp_small)
vis.gam(model_sp_large)

# if you're unsure of how many sp values you need:
m <- gam(y ~ te(x, z))
length(m$sp) # we need 2
```

# Question 2
Repeat Q1, but with Thin-plate splines

## Thin-plate spline with a penalty

```{r}
dat <- eg$data
y <- dat$y
x <- dat$x
z <- dat$z

thin_plate_penalty <- gam(y ~ s(x, z), data = dat)
vis.gam(thin_plate_penalty, main = "Thin-plate Spline with penalty")
```

## Thin-plate spline with no penalty

```{r}
thin_plate_no_penalty <- gam(y ~ s(x, z, fx = TRUE), data = dat)
vis.gam(thin_plate_no_penalty, main = "Thin-plate Spline without penalty")
```

## Thin-plate spline with choosing number of knots manually

```{r}
range(x)
range(z)
# Penalized with 5 knots in each direction
thin_plate_choose_knots <- gam(y ~ s(x, z,
                       k = 50), # now we set one k that calculates the joint basis functions
                data = dat)
vis.gam(thin_plate_choose_knots, main = "Thin-plate Spline by choosing number of knots")
```

## Thin-plate Spline choosing exact knot placement

You cannot specify exact knot placement for Thin-plate splines as it automatically calculates the knots that optimize smoothness.

## Thin-plate Spline with different number for smoothing parameter

```{r}
# Fix smoothing parameter to a small value (less smooth)
thin_plate_sp_small <- gam(y ~ s(x, z), data = dat, 
                      sp = 0.01) # now we only need one because thin plate splines penalizes the integral of squared derivatives over the whole 2D function, i.e. one joint value

# Fix smoothing parameter to a large value (very smooth)
thin_plate_sp_large <- gam(y ~ s(x, z), data = dat, sp = 10)

par(mfrow=c(1,2))
vis.gam(thin_plate_sp_small)
vis.gam(thin_plate_sp_large)

# if you're unsure of how many sp values you need:
m <- gam(y ~ s(x, z))
length(m$sp) # we only need 1
```

# Question 3

The question has 4 sub-steps:
1. Construct a tensor product basis using R functions for the marginal bases of x and z
2. Fit the model without the use of R functions
3. Write the code such that you can vary the number of knots
4. Plot the surface

We will write a function to do the following:
1. Take in values x and z as well as the number of knots for each as a vector
2. Construct the basis functions using the bs function for x and z individually
3. Combine the basis functions to for the tensor product basis
4. Return the tensor product basis

```{r}
create_tensor_product_basis <- function(x, z, num_knots){
  x_num_knots <- num_knots[1]
  z_num_knots <- num_knots[2]
  
  # determine the degrees of freedom
  x_df <- 3 + x_num_knots
  z_df <- 3 + z_num_knots
  
  # construct the marginal b-splines basis
  x_basis <- bs(x, df = x_df, degree = 3)
  z_basis <- bs(z, df = z_df, degree = 3)
  
  # combine to form tensor product
  num_cols <- ncol(x_basis)*ncol(z_basis)
  tensor_basis <- matrix(NA, nrow = length(x), ncol = num_cols)
  # take tensor product of all basis functions:
  cur_col <- 1
  for(i in 1:ncol(x_basis)){
    for(j in 1:ncol(z_basis)){
      tensor_basis[, cur_col] <- x_basis[,i]*z_basis[,j]
      cur_col <- cur_col + 1
    }
  }
  return(tensor_basis)
}
```

```{r}
# create the basis
tensor_prod_basis <- create_tensor_product_basis(x, z, num_knots = c(2,2))

# fit to the data using OLS
beta_hat <- solve(crossprod(tensor_prod_basis), crossprod(tensor_prod_basis, y))
y_fitted <- tensor_prod_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_tensor_product_basis(grid$x, grid$z, num_knots = c(2,2))  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Tensor Product Surface")
```

Repeat but with 10 knots for x and z.

```{r}
# create the basis
tensor_prod_basis <- create_tensor_product_basis(x, z, num_knots = c(10,10))

# fit to the data using OLS
beta_hat <- solve(crossprod(tensor_prod_basis), crossprod(tensor_prod_basis, y))
y_fitted <- tensor_prod_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_tensor_product_basis(grid$x, grid$z, num_knots = c(10,10))  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Tensor Product Surface")
```

# Question 4
## Thin-plate with random observations as knots

```{r}
create_thin_plate_random <- function(x, z, num_obs){
  dat <- cbind(x,z)
  # choose random index of observations
  index <- sample(1:length(x), num_obs)
  knots <- dat[index,]
  
  thin_plate_basis <- matrix(NA, ncol = (3+num_obs), nrow = length(x))
  
  thin_plate_basis[,1] <- 1 # column of ones for intercept
  thin_plate_basis[,2] <- x # x is column 2
  thin_plate_basis[,3] <- z # z is column 3
  
  # construct remaining basis functions
  ## calculate the distance:
  for(i in 1:num_obs){
    cur_knot <- knots[i,]
    r <- sqrt((x - cur_knot[1])^2 + (z - cur_knot[2])^2)  # Euclidean distance
    # Handle r = 0 carefully (log(0) is undefined)
    r[r == 0] <- 1e-10   # small number to avoid NaN
    basis <- r^2*log(r)
    thin_plate_basis[,3+i] <- basis
  }
  return(thin_plate_basis)
}

# create_thin_plate_random(x, z, 3)
```

```{r}
# apply the function
## create the basis
thin_plate_basis <- create_thin_plate_random(x, z, 3)


# fit to the data using OLS
beta_hat <- solve(crossprod(thin_plate_basis), crossprod(thin_plate_basis, y))
y_fitted <- thin_plate_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_thin_plate_random(grid$x, grid$z, 3)  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Thin-plate Surface")
```

Now refit is with more knots.

```{r}
# apply the function
## create the basis
thin_plate_basis <- create_thin_plate_random(x, z, 500)


# fit to the data using OLS
beta_hat <- solve(crossprod(thin_plate_basis), crossprod(thin_plate_basis, y))
y_fitted <- thin_plate_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_thin_plate_random(grid$x, grid$z, 500)  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Thin-plate Surface")
```

## Thin plate spline with equally spaced knots

```{r}
create_thin_plate_equally <- function(x, z, num_knots){
  dat <- cbind(x,z)
  # set equally spaced knots
  total_points <- num_knots + 2
  all_x_points <- seq(range(x)[1], range(x)[2], length.out = total_points)
  x_knots <- all_x_points[-c(1,total_points)] # remove first and last points to form interior knots
  all_z_points <- seq(range(z)[1], range(z)[2], length.out = total_points)
  z_knots <- all_z_points[-c(1,total_points)] # remove first and last points to form interior knots
  # now combine all combinations of these knots to form knots over the whole 2D plane
  knots <- expand.grid(x_knots, z_knots)

  
  thin_plate_basis <- matrix(NA, ncol = (3+ nrow(knots)), nrow = length(x))
  
  thin_plate_basis[,1] <- 1 # column of ones for intercept
  thin_plate_basis[,2] <- x # x is column 2
  thin_plate_basis[,3] <- z # z is column 3
  
  # construct remaining basis functions
  ## calculate the distance:
  for(i in 1:nrow(knots)){
    cur_knot <- as.numeric(knots[i,]) # a vector of knots
    r <- sqrt((x - cur_knot[1])^2 + (z - cur_knot[2])^2)  # Euclidean distance
    # Handle r = 0 carefully (log(0) is undefined)
    r[r == 0] <- 1e-10   # small number to avoid NaN
    basis <- r^2*log(r)
    thin_plate_basis[,3+i] <- basis
  }
  return(thin_plate_basis)
}

# create_thin_plate_equally(x, z, 4)
```

```{r}
# apply the function
## create the basis
thin_plate_basis <- create_thin_plate_equally(x, z, 5)

# fit to the data using OLS
beta_hat <- solve(crossprod(thin_plate_basis), crossprod(thin_plate_basis, y))
y_fitted <- thin_plate_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_thin_plate_equally(grid$x, grid$z, 5)  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Thin-plate Surface")
```

Now add more knots.

```{r}
# apply the function
## create the basis
thin_plate_basis <- create_thin_plate_equally(x, z, 10)

# fit to the data using OLS
beta_hat <- solve(crossprod(thin_plate_basis), crossprod(thin_plate_basis, y))
y_fitted <- thin_plate_basis%*%beta_hat

# now fit the surface
# Create regular grid
x_seq <- seq(min(x), max(x), length.out = 50)
z_seq <- seq(min(z), max(z), length.out = 50)
grid <- expand.grid(x = x_seq, z = z_seq)

# Create tensor basis for grid
grid_basis <- create_thin_plate_equally(grid$x, grid$z, 10)  

# Predict
grid_yhat <- grid_basis %*% beta_hat

# Reshape
z_matrix <- matrix(grid_yhat, nrow = length(x_seq), ncol = length(z_seq))

# 3D surface plot
persp(x_seq, z_seq, z_matrix,
      theta = 30, phi = 30, expand = 0.5,
      col = "lightblue", shade = 0.5,
      xlab = "x", ylab = "z", zlab = "Fitted y",
      main = "Fitted Thin-plate Surface")
```