---
title: "Tutorial 1: Splines"
author: "Andomei Smit: SMTAND051"
date: "14/04/2025"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: yes
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'H')
library(splines)
library(SemiPar)
```

# Question 1

## Simulate data

```{r}
rm(list = ls())
set.seed(1)
x_i <- runif(100) # x ~ U(0,1)
# define u_i
u_i <- 5 + sin(3*pi*(x_i-0.6))

# define y_i
y_i <- rnorm(100, mean = u_i, sd = 0.5)

plot(x_i, y_i)
```
Now, we will:
1. Define the basis functions
2. Define X
3. Plot the basis functions
4. Fit the basis functions to the data and plot

## Piecewise Constant

### Define the basis functions
Piecewise constant has one basis function: 1. This allows us to fit 3 different means in each of the sections. More specifically,

$$
B_j(x)=\begin{cases}
1, & \text{if } x \in [t_{j-1}, t_j) \\
0, & \text{otherwise}
\end{cases}
\quad \text{for } j = 1, 2
$$
and for the last knot we have

$$
B_3(x) =
\begin{cases}
1, & \text{if } x \in [t_{2}, t_3] \\
0, & \text{otherwise}
\end{cases}
$$

We also have
$$
t_1 = 0.33,\  t_2 = 0.66,\ t_3 =1
$$
although $t_3$ is not technically a knot, but rather an endpoint.

### Define X
```{r}
# X will be a 100x3 matrix. 
## for 0<=x<0.33 column 1 has ones, 0 otherwise
## for 0.33<=x<0.66 column 2 has ones, 0 otherwise
## for 0.66<=x<=1 column 3 has ones, 0 otherwise

# initialize design matrix
X <- matrix(0, nrow = length(x_i), ncol = 3)

# define intervals using indicator functions
X[,1] <- as.numeric(x_i < 0.33)
X[,2] <- as.numeric(x_i >= 0.33 & x_i < 0.66)
X[,3] <- as.numeric(x_i >= 0.66)
```

### Plot basis functions

```{r}
# Define the sequence of x values to evaluate the basis functions
x_vals <- seq(0, 1, length.out = 1000)

# Define the 3 basis functions as indicator functions
B1 <- as.numeric(x_vals < 0.33)
B2 <- as.numeric(x_vals >= 0.33 & x_vals < 0.66)
B3 <- as.numeric(x_vals >= 0.66)

# Plot
plot(x_vals, B1, type = "l", col = "red", lwd = 2,
     ylim = c(0, 1.2), ylab = "Basis Function Value", xlab = "x",
     main = "Piecewise Constant Spline Basis Functions")
lines(x_vals, B2, col = "blue", lwd = 2)
lines(x_vals, B3, col = "green", lwd = 2)

legend("topright", legend = c("B1: [0, 0.33)", "B2: [0.33, 0.66)", "B3: [0.66, 1]"),
       col = c("red", "blue", "green"), lwd = 2, 
       inset = 0.01,     # shrink distance from edge
       cex = 0.8)        # shrink text

```

### Fit the basis functions to the data and plot

```{r}
# Step 1: Simulate the data
## Done above

# Step 2: Construct the piecewise constant basis (design matrix)
X <- cbind(
  B1 = as.numeric(x_i < 0.33),
  B2 = as.numeric(x_i >= 0.33 & x_i < 0.66),
  B3 = as.numeric(x_i >= 0.66)
)

# Step 3: Fit the linear model without intercept (so basis acts directly)
fit <- lm(y_i ~ X - 1)  # "-1" removes the intercept

# Step 4: Plot the data and the fitted piecewise constant spline
# Sort x for nice plotting
sorted_index <- order(x_i)
x_sorted <- x_i[sorted_index]
y_sorted <- y_i[sorted_index]

# Predicted values using fitted model
y_pred <- fitted(fit)

# Step 5: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y", main = "Piecewise Constant Fit")
abline(v = c(0.33, 0.66), lty = 2, col = "blue")
# add the mean predicted lines:
lines(c(0, 0.33) , c(coef(fit)[1], coef(fit)[1]), col = "red", lwd = 2)
lines(c(0.33, 0.66) , c(coef(fit)[2], coef(fit)[2]), col = "red", lwd = 2)
lines(c(0.66, 1) , c(coef(fit)[3], coef(fit)[3]), col = "red", lwd = 2)
```

## Piecewise Linear
We now add a slope to each each piecewise function.

### Define the basis functions

$$
\begin{aligned}
B_1(x) &= \mathbb{1}(x < 0.33) \\
B_2(x) &= x \cdot \mathbb{1}(x < 0.33) \\
B_3(x) &= \mathbb{1}(0.33 \le x < 0.66) \\
B_4(x) &= x \cdot \mathbb{1}(0.33 \le x < 0.66) \\
B_5(x) &= \mathbb{1}(x \ge 0.66) \\
B_6(x) &= x \cdot \mathbb{1}(x \ge 0.66)
\end{aligned}
$$

### Define X

X now has a 1 in column 1 and 2 if x in [0, 0.33), it has a 1 in column 3 and 4 if x in [0.33, 0.66) and a 1 in column 5 and 6 if x in [0.66,1].

```{r}
x <- seq(0, 1, length.out = 1000)

B1 <- as.numeric(x < 0.33)
B2 <- x * (x < 0.33)

B3 <- as.numeric(x >= 0.33 & x < 0.66)
B4 <- x * (x >= 0.33 & x < 0.66)

B5 <- as.numeric(x >= 0.66)
B6 <- x * (x >= 0.66)

X <- cbind(B1, B2, B3, B4, B5, B6)
```


### Plot the basis functions

```{r}
plot(x, B1, type = "l", col = "black", ylim = c(0,1.1), lwd = 2,
     ylab = "Basis Function Value", xlab = "x", main = "Discontinuous Piecewise Linear Basis")
lines(x, B2, col = "red", lwd = 2)
lines(x, B3, col = "blue", lwd = 2)
lines(x, B4, col = "green", lwd = 2)
lines(x, B5, col = "orange", lwd = 2)
lines(x, B6, col = "purple", lwd = 2)

legend("topleft", 
       legend = c("B1 = 1 on [0,0.33)", "B2 = x on [0,0.33)", 
                  "B3 = 1 on [0.33,0.66)", "B4 = x on [0.33,0.66)", 
                  "B5 = 1 on [0.66,1]", "B6 = x on [0.66,1]"),
       col = c("black", "red", "blue", "green", "orange", "purple"), 
       lwd = 2, cex = 0.8)

```

### Fit the basis functions to the data and plot


```{r}
# Step 1: Simulate data
## done above

# Step 2: Define basis functions (discontinuous piecewise linear)
B1 <- as.numeric(x_i < 0.33)
B2 <- x_i * B1

B3 <- as.numeric(x_i >= 0.33 & x_i < 0.66)
B4 <- x_i * B3

B5 <- as.numeric(x_i >= 0.66)
B6 <- x_i * B5

# Step 3: Build design matrix
X <- cbind(B1, B2, B3, B4, B5, B6)

# Step 4: Fit the linear model
fit <- lm(y_i ~ X - 1)  # "-1" removes intercept; each region has its own intercept via B1, B3, B5

# Step 5: Predict over a dense grid for smooth plotting
x_plot <- seq(0, 1, length.out = 1000)

B1p <- as.numeric(x_plot < 0.33)
B2p <- x_plot * B1p

B3p <- as.numeric(x_plot >= 0.33 & x_plot < 0.66)
B4p <- x_plot * B3p

B5p <- as.numeric(x_plot >= 0.66)
B6p <- x_plot * B5p

Xp <- cbind(B1p, B2p, B3p, B4p, B5p, B6p)

# Predicted y values
y_fit <- Xp %*% coef(fit) # X*beta_hat

# Step 6: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y", main = "Discontinuous Piecewise Linear Fit")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(0.33, 0.66), col = "red", lty = 2)

```

## Piecewise linear (Continuous as the knots)
Now the basis functions need to be continuous at the knots, so the intercepts for each linear function needs to match up.

### Define the basis functions

We now drop the two basis functions related to the intercept of the second and third sections.

$$
\begin{aligned}
B_1(x) &= 1 \quad \text{(intercept)} \\
B_2(x) &= x \quad \text{(linear term)} \\
B_3(x) &= (x - 0.33)_+ = \max(0, x - 0.33) \\
B_4(x) &= (x - 0.66)_+ = \max(0, x - 0.66)
\end{aligned}
$$

### Define X

```{r}
# Define x values over the domain
x <- seq(0, 1, length.out = 1000)

# Define the knots
knot1 <- 0.33
knot2 <- 0.66

# Construct the truncated power basis (degree 1 spline, continuous at knots)
B1 <- rep(1, length(x))           # Intercept for all x
B2 <- x                           # Linear term for all x
B3 <- pmax(0, x - knot1)          # (x - 0.33)_+ for x>=0.33
B4 <- pmax(0, x - knot2)          # (x - 0.66)_+ for x>=0.66

# Combine into a design matrix
X <- cbind(B1, B2, B3, B4)
```

### Plot the basis functions

```{r}
# Plot all basis functions
plot(x, B1, type = "l", lwd = 2, col = "black", ylim = c(0, 1.1),
     xlab = "x", ylab = "Basis Function Value", main = "Continuous Piecewise Linear Basis Functions")
lines(x, B2, col = "blue", lwd = 2)
lines(x, B3, col = "red", lwd = 2)
lines(x, B4, col = "green", lwd = 2)

# Add a legend
legend("topleft",
       legend = c("B1 = 1", "B2 = x", "B3 = (x - 0.33)+", "B4 = (x - 0.66)+"),
       col = c("black", "blue", "red", "green"),
       lwd = 2, cex = 0.8)
```

### Fit the basis functions to the data and plot

We now want to fit:

$$
y_i = \beta_0\  +\  \beta_1x_i\ +\ \beta_2(x_i-0.33)_+\ +\ \beta_3(x_i-0.66)_+
$$

```{r}
# Step 1: Simulate the data
## done above

# Step 2: Define the basis functions for the truncated power basis
knot1 <- 0.33
knot2 <- 0.66

B1 <- rep(1, length(x_i))           # Intercept
B2 <- x_i                           # x
B3 <- pmax(0, x_i - knot1)          # (x - 0.33)_+
B4 <- pmax(0, x_i - knot2)          # (x - 0.66)_+

X <- cbind(B1, B2, B3, B4)

# Step 3: Fit the linear model
fit <- lm(y_i ~ X - 1)  # "-1" removes automatic intercept, since B1 handles it

# Optional: check model summary
summary(fit)
```

```{r}
# Step 4: Predict over a smooth grid
x_plot <- seq(0, 1, length.out = 1000)

B1p <- rep(1, length(x_plot))
B2p <- x_plot
B3p <- pmax(0, x_plot - knot1)
B4p <- pmax(0, x_plot - knot2)

Xp <- cbind(B1p, B2p, B3p, B4p)
y_fit <- Xp %*% coef(fit) # y_hat = XB_hat

# Step 5: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y", main = "Continuous Piecewise Linear Fit")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(knot1, knot2), col = "red", lty = 2)
```


## Piecewise cubic polynomial (discontinuous)
We define a cubic polynomial for each interval.

### Define the basis functions


$$
\begin{aligned}
B_1(x) &= \mathbb{1}(x < 0.33), \\ 
B_2(x) &= x \cdot \mathbb{1}(x < 0.33), \\ 
B_3(x) &= x^2 \cdot \mathbb{1}(x < 0.33), \\
B_4(x) &= x^3 \cdot \mathbb{1}(x < 0.33), \\[1em]
B_5(x) &= \mathbb{1}(0.33 \le x < 0.66), \\
B_6(x) &= x \cdot \mathbb{1}(0.33 \le x < 0.66), \\
B_7(x) &= x^2 \cdot \mathbb{1}(0.33 \le x < 0.66), \\
B_8(x) &= x^3 \cdot \mathbb{1}(0.33 \le x < 0.66), \\[1em]
B_9(x) &= \mathbb{1}(x \ge 0.66), \\ 
B_{10}(x) &= x \cdot \mathbb{1}(x \ge 0.66), \\ 
B_{11}(x) &= x^2 \cdot \mathbb{1}(x \ge 0.66), \\
B_{12}(x) &= x^3 \cdot \mathbb{1}(x \ge 0.66)
\end{aligned}
$$
This spline will have discontinuities at the knots.

### Define X

```{r}
# Interval 1: x in [0, 0.33)
B1 <- as.numeric(x_i < knot1)
B2 <- x_i * B1
B3 <- x_i^2 * B1
B4 <- x_i^3 * B1

# Interval 2: x in [0.33, 0.66)
B5 <- as.numeric(x_i >= knot1 & x_i < knot2)
B6 <- x_i * B5
B7 <- x_i^2 * B5
B8 <- x_i^3 * B5

# Interval 3: x in [0.66, 1]
B9  <- as.numeric(x_i >= knot2) # define an indicator for the third interval
B10 <- x_i * B9
B11 <- x_i^2 * B9
B12 <- x_i^3 * B9

# Combine into design matrix
X <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8,
           B9, B10, B11, B12)

# Optional: check dimensions and column names
colnames(X) <- paste0("B", 1:12)
dim(X)
```

### Plot the basis functions

```{r}
# Define plotting grid
x <- seq(0, 1, length.out = 1000)

# Define basis functions on the grid
# Interval 1: x in [0, 0.33)
B1 <- as.numeric(x < knot1)
B2 <- x * B1
B3 <- x^2 * B1
B4 <- x^3 * B1

# Interval 2: x in [0.33, 0.66)
B5 <- as.numeric(x >= knot1 & x < knot2)
B6 <- x * B5
B7 <- x^2 * B5
B8 <- x^3 * B5

# Interval 3: x in [0.66, 1]
B9  <- as.numeric(x >= knot2) # define an indicator for the third interval
B10 <- x * B9
B11 <- x^2 * B9
B12 <- x^3 * B9

# Combine for plotting
basis_matrix <- cbind(B1, B2, B3, B4, B5, B6, B7, B8, B9, B10, B11, B12)

# Plot
matplot(x, basis_matrix, type = "l", lty = 1, lwd = 2,
        col = rainbow(6), ylab = "Basis Function Value", xlab = "x",
        main = "Piecewise Cubic Basis Functions (Continuous at Knots)")

legend("topright", legend = paste0("B", 1:6),
       col = rainbow(6), lwd = 2, cex = 0.8)
```

### Fit the basis functions to the data and plot

```{r}

# Step 3: Create design matrix for x_i
## Use from above in step 2

# Step 4: Fit the linear model
fit <- lm(y_i ~ X - 1)  # remove intercept as B1 handles it

# Step 5: Predict on a dense grid
## use the betas defined above to plot the basis

y_fit <- basis_matrix %*% coef(fit)

# Step 6: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y",
     main = "Piecewise Cubic Spline Fit (Continuous at Knots)")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(knot1, knot2), col = "red", lty = 2)
```

## Piecewise cubic polynomial, continuous at the knots
We now fit three different cubic polynomials with no constraints on the knots.

### Define the basis functions

We can define $f(x)$ piecewise as:
$$
f(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \gamma_1 (x - \xi_1)_+ + \gamma_2 (x - \xi_1)^2_+ + \gamma_3 (x - \xi_1)^3_+ \\ + \gamma_4 (x - \xi_2)_+ + \gamma_5 (x - \xi_2)^2_+ + \gamma_6 (x - \xi_2)^3_+
$$

### Define X

```{r}
# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- pmax(0, x_i - knot1) # linear
B6 <- (max(0, x_i - knot1))^2 # quadratic
B7 <- (max(0, x_i - knot1))^3 # cubic


B8 <- pmax(0, x_i - knot2) # linear
B9 <- (max(0, x_i - knot2))^2 # quadratic
B10 <- (max(0, x_i - knot2))^3 # cubic

# Combine into design matrix
X <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8,
           B9, B10)

# Optional: check dimensions and column names
colnames(X) <- paste0("B", 1:10)
dim(X)
```

### Plot the basis functions

```{r}
# Define a fine grid for plotting
x <- seq(0, 1, length.out = 1000)

# Global polynomial
B1 <- rep(1, length(x))
B2 <- x 
B3 <- x^2 
B4 <- x^3 

# truncated power bases:
B5 <- pmax(0, x - knot1) # linear
B6 <- (pmax(0, x - knot1))^2 # quadratic
B7 <- (pmax(0, x - knot1))^3 # cubic


B8 <- pmax(0, x - knot2) # linear
B9 <- (pmax(0, x - knot2))^2 # quadratic
B10 <- (pmax(0, x - knot2))^3 # cubic

# Combine into design matrix
basis_matrix <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8,
           B9, B10)

# Plot all basis functions
matplot(x, basis_matrix, type = "l", lty = 1, lwd = 2,
        col = rainbow(10), ylab = "Basis Value", xlab = "x",
        main = "Piecewise Cubic Polynomial Basis (Continuous at the knots)")

legend("topright", legend = paste0("B", 1:10), 
       col = rainbow(10), lwd = 2, ncol = 2, cex = 0.75)
```

### Fit the basis functions to the data and plot

```{r}
# Step 1: Simulate the data
## done above

# Step 2: Create the design matrix (same as before)

# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- pmax(0, x_i - knot1) # linear
B6 <- (pmax(0, x_i - knot1))^2 # quadratic
B7 <- (pmax(0, x_i - knot1))^3 # cubic

B8 <- pmax(0, x_i - knot2) # linear
B9 <- (pmax(0, x_i - knot2))^2 # quadratic
B10 <- (pmax(0, x_i - knot2))^3 # cubic

X <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8,
           B9, B10)

# Step 3: Fit the model
fit <- lm(y_i ~ X - 1)  # Remove intercept since B1 already covers it

# Step 4: Predict on a fine grid
x_plot <- seq(0, 1, length.out = 1000)

# Basis functions for x_plot
# Global polynomial
B1p <- rep(1, length(x_plot))
B2p <- x_plot 
B3p <- x_plot^2 
B4p <- x_plot^3 

# truncated power bases:
B5p <- pmax(0, x_plot - knot1) # linear
B6p <- (pmax(0, x_plot - knot1))^2 # quadratic
B7p <- (pmax(0, x_plot - knot1))^3 # cubic

B8p <- pmax(0, x_plot - knot2) # linear
B9p <- (pmax(0, x_plot - knot2))^2 # quadratic
B10p <- (pmax(0, x_plot - knot2))^3 # cubic

Xp <- cbind(B1p, B2p, B3p, B4p,
            B5p, B6p, B7p, B8p,
            B9p, B10p)

y_fit <- Xp %*% coef(fit)

# Step 5: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y",
     main = "Piecewise Cubic Polynomia, continuous at the knots")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(knot1, knot2), col = "red", lty = 2)
```


## Piecewise cubic polynomial continuous up to first derivative

### Define the basis functions
Now we drop the linear truncated power basis to make it continuous up to first derivative.

We can define $f(x)$ piecewise as:
$$
f(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3  + \gamma_1 (x - \xi_1)^2_+ + \gamma_2 (x - \xi_1)^3_+ \\ + \gamma_3 (x - \xi_2)^2_+ + \gamma_4 (x - \xi_2)^3_+
$$
### Define X

```{r}
# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- (max(0, x_i - knot1))^2 # quadratic
B6 <- (max(0, x_i - knot1))^3 # cubic

B7 <- (max(0, x_i - knot2))^2 # quadratic
B8 <- (max(0, x_i - knot2))^3 # cubic

# Combine into design matrix
X <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8)

# Optional: check dimensions and column names
colnames(X) <- paste0("B", 1:8)
dim(X)
```

### Plot the basis functions

```{r}
# Define a fine grid for plotting
x <- seq(0, 1, length.out = 1000)

# Global polynomial
B1 <- rep(1, length(x))
B2 <- x 
B3 <- x^2 
B4 <- x^3 

# truncated power bases:
B5 <- (pmax(0, x - knot1))^2 # quadratic
B6 <- (pmax(0, x - knot1))^3 # cubic

B7 <- (pmax(0, x - knot2))^2 # quadratic
B8 <- (pmax(0, x - knot2))^3 # cubic

# Combine into design matrix
basis_matrix <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8)

# Plot all basis functions
matplot(x, basis_matrix, type = "l", lty = 1, lwd = 2,
        col = rainbow(10), ylab = "Basis Value", xlab = "x",
        main = "Piecewise Cubic Polynomial Basis (Continuous at the knots)")

legend("topright", legend = paste0("B", 1:10), 
       col = rainbow(10), lwd = 2, ncol = 2, cex = 0.75)
```

### Fit the basis functions to the data and plot

```{r}
# Step 1: Simulate the data
## done above

# Step 2: Create the design matrix (same as before)

# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- (pmax(0, x_i - knot1))^2 # quadratic
B6 <- (pmax(0, x_i - knot1))^3 # cubic

B7 <- (pmax(0, x_i - knot2))^2 # quadratic
B8 <- (pmax(0, x_i - knot2))^3 # cubic

X <- cbind(B1, B2, B3, B4,
           B5, B6, B7, B8)

# Step 3: Fit the model
fit <- lm(y_i ~ X - 1)  # Remove intercept since B1 already covers it

# Step 4: Predict on a fine grid
x_plot <- seq(0, 1, length.out = 1000)

# Basis functions for x_plot
# Global polynomial
B1p <- rep(1, length(x_plot))
B2p <- x_plot 
B3p <- x_plot^2 
B4p <- x_plot^3 

# truncated power bases:
B5p <- (pmax(0, x_plot - knot1))^2 # quadratic
B6p <- (pmax(0, x_plot - knot1))^3 # cubic

B7p <- (pmax(0, x_plot - knot2))^2 # quadratic
B8p <- (pmax(0, x_plot - knot2))^3 # cubic

Xp <- cbind(B1p, B2p, B3p, B4p,
            B5p, B6p, B7p, B8p)

y_fit <- Xp %*% coef(fit)

# Step 5: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y",
     main = "Piecewise Cubic Polynomia, continuous at the knots")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(knot1, knot2), col = "red", lty = 2)
```

## Piecewise cubic polynomial continuous up to second derivative
Now we drop the quadratic truncated power basis too.

### Define the basis functions

We can define $f(x)$ piecewise as:
$$
f(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3  + \gamma_1 (x - \xi_1)^3_+ + \gamma_2 (x - \xi_2)^3_+
$$

### Define X

```{r}
# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- (max(0, x_i - knot1))^3 # cubic

B6 <- (max(0, x_i - knot2))^3 # cubic

# Combine into design matrix
X <- cbind(B1, B2, B3, B4,
           B5, B6)

# Optional: check dimensions and column names
colnames(X) <- paste0("B", 1:6)
dim(X)
```

### Plot the basis functions

```{r}
# Define a fine grid for plotting
x <- seq(0, 1, length.out = 1000)

# Global polynomial
B1 <- rep(1, length(x))
B2 <- x 
B3 <- x^2 
B4 <- x^3 

# truncated power bases:
B5 <- (pmax(0, x - knot1))^3 # cubic
B6 <- (pmax(0, x - knot2))^3 # cubic

# Combine into design matrix
basis_matrix <- cbind(B1, B2, B3, B4,
           B5, B6)

# Plot all basis functions
matplot(x, basis_matrix, type = "l", lty = 1, lwd = 2,
        col = rainbow(6), ylab = "Basis Value", xlab = "x",
        main = "Piecewise Cubic Polynomial Basis (Continuous at the knots)")

legend("topright", legend = paste0("B", 1:6), 
       col = rainbow(6), lwd = 2, ncol = 2, cex = 0.75)
```

### Fit the basis functions to the data and plot

```{r}
# Step 1: Simulate the data
## done above

# Step 2: Create the design matrix (same as before)

# Global polynomial
B1 <- rep(1, length(x_i))
B2 <- x_i 
B3 <- x_i^2 
B4 <- x_i^3 

# truncated power bases:
B5 <- (pmax(0, x_i - knot1))^3 # cubic
B6 <- (pmax(0, x_i - knot2))^3 # cubic

X <- cbind(B1, B2, B3, B4,
           B5, B6)

# Step 3: Fit the model
fit <- lm(y_i ~ X - 1)  # Remove intercept since B1 already covers it

# Step 4: Predict on a fine grid
x_plot <- seq(0, 1, length.out = 1000)

# Basis functions for x_plot
# Global polynomial
B1p <- rep(1, length(x_plot))
B2p <- x_plot 
B3p <- x_plot^2 
B4p <- x_plot^3 

# truncated power bases:
B5p <- (pmax(0, x_plot - knot1))^3 # cubic
B6p <- (pmax(0, x_plot - knot2))^3 # cubic

Xp <- cbind(B1p, B2p, B3p, B4p,
            B5p, B6p)

y_fit <- Xp %*% coef(fit)

# Step 5: Plot
plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y",
     main = "Piecewise Cubic Polynomia, continuous at the knots")
lines(x_plot, y_fit, col = "blue", lwd = 2)
abline(v = c(knot1, knot2), col = "red", lty = 2)
```

## Natural spline

We will use the ns() function from the splines package. Note that we will assume the boundary knots are at 0 and 1, not at the two knots (0.33 and 0.66).

### Define the basis function

The ns() package defines the basis functions differently to ensure linearity past the boundary knots, but in general, from first principles, the basis functions are:

$$
f(x) = \beta_0 + \beta_1 x + \sum_{k=1}^{K} \theta_k \cdot N_k(x)
$$

$$
N_k(x) = d_k(x) - d_K(x)
$$

$$
d_k(x) = \frac{(x - \xi_k)^3_+ - (x - \xi_K)^3_+}{\xi_K - \xi_k}
$$

$$
(x - \xi)^3_+ = 
\begin{cases}
(x - \xi)^3 & \text{if } x > \xi \\\\
0 & \text{otherwise}
\end{cases}
$$

Plotting the ns() basis functions:

```{r}
x_grid <- seq(0, 1, length.out = 200)
B <- ns(x_grid, knots = c(0.33, 0.66))

matplot(x_grid, B, type = "l", lty = 1, lwd = 2,
        col = 1:ncol(B), ylab = "Basis Function Value",
        main = "Natural Spline Basis Functions")
legend("topright", legend = paste0("B", 1:ncol(B)), col = 1:ncol(B), lty = 1)

```

### Define X

```{r}
ns(x_i, knots = c(0.33, 0.66)) # note that they do not include the intercept
```


### Plot the basis functions

### Fit the basis functions to the data

```{r}
# Fit model using natural spline with specified knots
model_ns <- lm(y_i ~ ns(x_i, knots = c(0.33, 0.66))) # note we do not subtract 1 here
```

```{r}
# Generate a smooth sequence of x values for plotting
x_grid <- seq(0, 1, length.out = 1000)

# Predict y values using the fitted spline model
# Extract boundary knots from training data
bknots <- range(x_i)

# Generate basis on x_grid with same internal and boundary knots
B_grid <- ns(x_grid, knots = c(0.33, 0.66), Boundary.knots = bknots)

# Compute predicted values manually
y_pred <- cbind(1, B_grid) %*% coef(model_ns)

# Plot the data and the fitted spline
plot(x_i, y_i, pch = 16, col = "grey", main = "Natural Spline Fit", xlab = "x", ylab = "y")
lines(x_grid, y_pred, col = "blue", lwd = 2)
abline(v = c(0.33, 0.66), col = "red", lty = 2)  # show knots

```

## B-sline basis function

Note again here we use the bs() function to define the basis. These basis functions are non-zero only over a short range of the data each. Plot these:

```{r}
B <- bs(x_grid, knots = c(0.33, 0.66))
matplot(x_grid, B, type = "l", lty = 1, lwd = 2,
        col = 1:ncol(B), ylab = "Basis Function Value",
        main = "B-Spline Basis Functions")
legend("topright", legend = paste0("B", 1:ncol(B)), col = 1:ncol(B), lty = 1)
```


### Define X

```{r}
bs(x_i, knots = c(0.33, 0.66))
```


### Fit the basis functions to the data

```{r}
# Fit model using B-spline basis
model_bs <- lm(y_i ~ bs(x_i, knots = c(0.33, 0.66))) # default: creates B-splines of degree 3
```

Plot these:
```{r}
# Prediction grid
x_grid <- seq(0, 1, length.out = 1000)
y_pred <- cbind(1, bs(x_grid, knots = c(0.33, 0.66))) %*% coef(model_bs)

# Plot
plot(x_i, y_i, pch = 16, col = "grey", main = "B-Spline Fit", xlab = "x", ylab = "y")
lines(x_grid, y_pred, col = "darkgreen", lwd = 2)
abline(v = c(0.33, 0.66), col = "red", lty = 2)
```

## Adding confidence bands to the B-spline

We have: 
$$
\text{var}(\hat{f}(x)) = X \Sigma X^T
$$

where

$$
\Sigma_{\beta} = \hat{\sigma}^2 (X^TX)^{-1}
$$

$$
\hat{\sigma}^2 = \frac{1}{n - p} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
$$
Thus we can calculate it here:
```{r}
# Create design matrix (basis functions) at new x values
X_grid <- cbind(1, bs(x_grid, knots = c(0.33, 0.66), Boundary.knots = range(x)))


# Residual variance
residuals <- resid(model_bs)
n <- length(y_i)
p <- length(coef(model_bs))
sigma_hat2 <- sum(residuals^2) / (n - p)

# Design matrix from training data (same structure as above)
X_train <- model.matrix(model_bs)

# covariance matrix of coefficients
Sigma_beta <- sigma_hat2 * solve(t(X_train) %*% X_train)

# pointwise se of predicted values
se_f_hat <- sqrt(rowSums((X_grid %*% Sigma_beta) * X_grid))  # sqrt of diagonal of var(f_hat)

# now plot:
fitted_vals <- X_grid %*% coef(model_bs)
upper <- fitted_vals + 1.96 * se_f_hat
lower <- fitted_vals - 1.96 * se_f_hat

plot(x_i, y_i, pch = 16, col = "grey", xlab = "x", ylab = "y", main = "B-spline with 95% CI")
lines(x_grid, fitted_vals, col = "blue", lwd = 2)
lines(x_grid, upper, col = "blue", lty = 2)
lines(x_grid, lower, col = "blue", lty = 2)
polygon(c(x_grid, rev(x_grid)), c(upper, rev(lower)), col = rgb(0, 0, 1, 0.1), border = NA)

```

# Question 2
Instruction: Write an R function to create a cubic spline basis based on a truncated power basis. Construct the basis matrix ð‘‹. Fit a cubic spline to the LIDAR data. Add the fitted spline curve to a plot with the observations. The LIDAR data are available in the SemiPar library.

## Load data
```{r}
rm(list=ls())
data("lidar")
x <- lidar$range
y <- lidar$logratio

plot(x, y)
```

## Write the R function

Given data, return the design matrix, X, that uses the cubic spline basis with truncated power bases. For now, we will assume the function is specific to the LIDAR data to determine the knots.

```{r}
cubic_spline_basis <- function(x, knots = c(550, 620)){
  b1 <- rep(1, length(x)) # global intercept
  b2 <- x # global linear
  b3 <- x^2 # global quadratic
  b4 <- x^3 # global cubic
  b5 <- (pmax(0, x-knots[1]))^3 # truncated cubic power basis for knot 1
  b6 <- (pmax(0, x-knots[2]))^3 # truncated cubic power basis for knot 2
  
  basis_matrix <- cbind(b1, b2, b3, b4, b5, b6)
  return(basis_matrix)
}

# test that it works: the basis matrix is:
cubic_basis <- cubic_spline_basis(x)

# now fit the cubic spline to the LIDAR data
model_cubic <- lm(y ~ cubic_basis - 1)

# create many x for a nice plot
x_plot <- seq(from = min(x), to = max(x), length.out=1000)
basis_plot <- cubic_spline_basis(x_plot)
y_plot <- basis_plot %*% coef(model_cubic)

# plot
plot(x, y, pch = 16, col = "grey", xlab = "x", ylab = "y",
     main = "Cubic Polynomial with truncated power basis on LIDAR data")
lines(x_plot, y_plot, col = "blue", lwd = 2)
abline(v = c(550, 620), col = "red", lty = 2)
```

# Question 3
Instructions: Construct the B-spline basis functions (see ESL Appendix pg. 186): Haar basis, linear, quadratic, cubic B-spline basis functions. x between -1, 1. Create basis matrix X. Plot the basis functions.


The following function will fit an order M B-splines basis to any sequence x with a set position for the knots.

```{r}
construct_bspline_basis <- function(x, knots, order) {
  # Step 1: Augment knot sequence
  augmented_knots <- c(rep(min(knots), order), knots, rep(max(knots), order))
  n_basis <- length(augmented_knots) - order - 1 # total number of b-spline bases we can construct
  
  # Step 2: Initialize Haar basis (order 1)
  basis_list <- list()
  B <- matrix(0, nrow = length(x), ncol = n_basis)
  for (i in 1:n_basis) {
  left <- augmented_knots[i]
  right <- augmented_knots[i + 1]
  
  if (i < n_basis) {
    B[, i] <- as.numeric(x >= left & x < right)
  } else {
    B[, i] <- as.numeric(x >= left & x <= right)  # include right endpoint for final basis
    }
  }

  basis_list[[1]] <- B
  
  if(order>1){ # only continue if order > 1
  # Step 3: Recursively build up to desired order
  for (m in 2:order) {
    prev_B <- basis_list[[m - 1]]
    prev_B <- cbind(prev_B, 0) # pad with zero for endpoint
    new_B <- matrix(0, nrow = length(x), ncol = ncol(prev_B) - 1)
    for (i in 1:(ncol(prev_B) - 1)) {
      denom1 <- augmented_knots[i + m - 1] - augmented_knots[i]
      denom2 <- augmented_knots[i + m] - augmented_knots[i + 1]
      
      T1 <- if (denom1 == 0) 0 else (x - augmented_knots[i]) / denom1 * prev_B[, i]
      T2 <- if (denom2 == 0) 0 else (augmented_knots[i + m] - x) / denom2 * prev_B[, i + 1]
      
      new_B[, i] <- T1 + T2
      }
    basis_list[[m]] <- new_B
    }
  }
  
  return(basis_list[[order]])
}
```

```{r}
# define x and the knots
x <- seq(-1, 1, length.out = 1000)
knots <- seq(-1, 1, by = 0.1)
```

## Haar (order 1 B-splines basis)

```{r}
B_cubic <- construct_bspline_basis(x, knots, order = 1)

matplot(x, B_cubic, type = "l", lty = 1, col = rainbow(ncol(B_cubic)),
        ylab = "Basis Value", main = "Haar B-spline Basis Functions (Order 1)")
```

## B-spline of order 2

```{r}
B_cubic <- construct_bspline_basis(x, knots, order = 2)

matplot(x, B_cubic, type = "l", lty = 1, col = rainbow(ncol(B_cubic)),
        ylab = "Basis Value", main = "B-spline Basis Functions (Order 2)")
```

## B-spline of order 3

```{r}
B_cubic <- construct_bspline_basis(x, knots, order = 3)

matplot(x, B_cubic, type = "l", lty = 1, col = rainbow(ncol(B_cubic)),
        ylab = "Basis Value", main = "B-spline Basis Functions (Order 3)")
```

## B-spline of order 4

```{r}
B_cubic <- construct_bspline_basis(x, knots, order = 4)

matplot(x, B_cubic, type = "l", lty = 1, col = rainbow(ncol(B_cubic)),
        ylab = "Basis Value", main = "B-spline Basis Functions (Order 4)")
```